---
title: "Draft: Results"
editor: 
  markdown: 
    wrap: sentence
    
bibliography: references.bib
---

# Data & Methods

The analysis uses publicly available bridge inspection and transportation datasets from the **Federal Highway Administration** and summary reports from the **Nebraska Department of Transportation**.
These datasets contain information on bridge conditions, construction characteristics, and traffic volumes for bridges across the United States.
The Nebraska bridge inventory dataset for 2025 was used to evaluate bridge conditions in **Douglas County**, which includes the city of **Omaha**.
The purpose of the analysis is to estimate the potential risk of bridge collapse by examining structural condition ratings and transportation usage.

## Data Sources

-   **National Bridge Inventory (NBI)**: A comprehensive dataset maintained by the @fhwaElement2025 that includes inspection results, structural condition ratings, bridge dimensions, construction year, and traffic statistics for bridges in the United States.^[The Nebraska NBI extract was downloaded from the FHWA bridge data portal at <https://www.fhwa.dot.gov/bridge/nbi/ascii.cfm>.]
-   **Nebraska Department of Transportation Bridge Reports**: Provides summary information on bridge conditions across Nebraska, including counts of structurally deficient or functionally obsolete bridges and maintenance planning data [@BridgeNDOT]. ^[Reports are publicly available at <https://dot.nebraska.gov/media/4efjqkfa/bridge-condition-locally-owned.pdf>.]

## Data Features

The datasets include several variables relevant to evaluating bridge safety and infrastructure risk:

-   **Structural Condition Ratings**: Includes inspection scores for deck condition `DECK_COND_058`, superstructure condition `SUPERSTRUCTURE_COND_059`, and substructure condition `SUBSTRUCTURE_COND_060`.
    These ratings range from 0 to 9, where higher values represent better structural condition.

-   **Traffic Volume**: Average Daily Traffic `ADT_029` estimates the number of vehicles crossing each bridge per day, indicating the level of transportation demand.

-   **Bridge Age**: Construction year `YEAR_BUILT_027` allows calculation of bridge age, which may influence structural deterioration over time.

-   **Location Information**: Geographic identifiers such as state code and county code allow bridges in Douglas County to be isolated for analysis.

-   **Bridge Dimensions and Characteristics**: Additional variables such as bridge length, roadway width, and number of spans describe the structural characteristics of each bridge


### Missing Data 

Prior to filteriing to Douglas County, a missing data assessment was conducted on the key structural condition variables. Missingess was observed primarily in the three condition rating fields (`DECK_COND_58`, `SUPERSTRUCTURE_COND_059`, `SUBSTRUCTURE_COND_060`) and in `ADT_029`. Bridges with missing values in any of these variables were excluded from analysis using `drop_na()`, as these fields are essential for risk classification. It is worth noting that missingness in condition ratings may not be random, smaller or lower-priority structures are sometimes exempt from full NBI inspection requirements, which could mean the excluded bridges are systematically different from those retained. This is a limitation of the analysis.

## Data Collection and Preparation

The Nebraska National Bridge Inventory dataset was downloaded in delimited text format and converted into a CSV file for analysis.The dataset was then imported into the R programming environment. Package versions and system information are documented in @sec-appendix-session

To focus on the study area, bridges located in Douglas County were selected using the county code variable (`COUNTY_CODE_003 == 55`).
Data cleaning steps were performed to prepare the dataset for statistical analysis.
Several inspection variables were initially stored as character values and were converted to numeric format in order to calculate summary statistics.
Missing or non-numeric values were treated as missing observations and excluded.

Additional variables were created to support analysis.
Bridge age was calculated by subtracting the construction year from the most current year (2025).
This variable helps identify older infrastructure that may require increased maintenance or monitoring.

## Data Classification

To evaluate structural risk, bridges were categorized according to their inspection ratings.

-   **Higher-risk bridges**: Bridges with structural condition ratings of 4 or lower in the deck, superstructure, or substructure categories, indicating poor structural condition or advanced deterioration.

-   **Moderate-condition bridges**: Bridges with ratings between 5 and 6, indicating fair structural condition.

-   **Good-condition bridges**: Bridges with ratings of 7 or higher, representing satisfactory or good structural performance.

This classification aligns with FHWA conventions for identifying structurally deficient bridges and allows the analysis to prioritize structure most in need of intervention.

## R packages and Software

The analysis was conducted in R. Package versions are recorded in @sec-appendix-session. Key packages used include:

-   **readr:** Used to import the bridge inventory dataset into R [@wickhamReadrReadRectangular2026].

-   **tidyverse:** Used for data manipulation, filtering, and summarization [@wickhamTidyverseEasilyInstall2023].

-   **dplyr:** Used to calculate descriptive statistics and identify high-risk bridges [@wickhamDplyrGrammarData2026].

-   **ggplot2:** Used to create visualizations of bridge condition ratings, traffic levels, and bridge age distributions [@wickhamGgplot2CreateElegant2026].

-   **knitr:** Used to generate the final report and integrate code with written analysis [@KnitrYihuiXie].

- **scales**


## Analytical Approach

The analysis proceeded in three stages. First, descriptive statistics were calculated to characterize the full sample of Douglas County bridges on key variables: structural condition ratings, bridge age, and traffic volume. Second, a condition rating analysis identified bridges with ratings at or below 4 in any structural component, flagging them as higher-risk. Third, a combined risk assessment cross-tabulated structural condition against traffic exposure (ADT), since bridges that are both structurally compromised and heavily used represent the greatest public safety risk — high likelihood of failure combined with high consequence. Spearman correlation tests were used to assess the monotonic relationships between age, condition, and traffic, given the ordinal nature of the condition rating variables.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(readr)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(knitr)
library(scales)

```

```{r,echo=FALSE, warning=FALSE}
inventory <- read.csv("data/national_bridge_inventory_ne.csv")

douglas_bridges <- inventory %>%
  select(
    COUNTY_CODE_003,
    YEAR_BUILT_027,
    YEAR_RECONSTRUCTED_106,
    ADT_029,
    FUNCTIONAL_CLASS_026,
    DECK_COND_058,
    SUPERSTRUCTURE_COND_059,
    SUBSTRUCTURE_COND_060,
    LOWEST_RATING,
    BRIDGE_CONDITION,
    STRUCTURE_LEN_MT_049,
    DECK_AREA
  ) %>%
  filter(COUNTY_CODE_003 == 55) %>%
  mutate(
    YEAR_BUILT_027 = as.numeric(YEAR_BUILT_027),
    ADT_029 = as.numeric(ADT_029),
    DECK_COND_058 = as.numeric(DECK_COND_058),
    SUPERSTRUCTURE_COND_059 = as.numeric(SUPERSTRUCTURE_COND_059),
    SUBSTRUCTURE_COND_060 = as.numeric(SUBSTRUCTURE_COND_060),
    bridge_age = 2026 - YEAR_BUILT_027
  ) %>%
  drop_na(ADT_029, bridge_age, DECK_COND_058, SUPERSTRUCTURE_COND_059, SUBSTRUCTURE_COND_060)

douglas_bridges <- douglas_bridges %>%
  mutate(
    risk_group = ifelse(
      DECK_COND_058 <= 4 |
      SUPERSTRUCTURE_COND_059 <= 4 |
      SUBSTRUCTURE_COND_060 <= 4,
      "High Risk",
      "Lower Risk"
    )
  )
```


```{r, echo=FALSE}
n_bridges <- nrow(douglas_bridges)

adt_min <- min(douglas_bridges$ADT_029, na.rm = TRUE)
adt_max <- max(douglas_bridges$ADT_029, na.rm = TRUE)
adt_median <- median(douglas_bridges$ADT_029, na.rm = TRUE)
adt_mean <- mean(douglas_bridges$ADT_029, na.rm = TRUE)

age_min <- min(douglas_bridges$bridge_age, na.rm = TRUE)
age_max <- max(douglas_bridges$bridge_age, na.rm = TRUE)
age_median <- median(douglas_bridges$bridge_age, na.rm = TRUE)
age_mean <- mean(douglas_bridges$bridge_age, na.rm = TRUE)

deck_mean <- mean(douglas_bridges$DECK_COND_058, na.rm = TRUE)
super_mean <- mean(douglas_bridges$SUPERSTRUCTURE_COND_059, na.rm = TRUE)
sub_mean <- mean(douglas_bridges$SUBSTRUCTURE_COND_060, na.rm = TRUE)

```
# Results

## Study Sample Characteristics

Bridge inspection data from the Federal Highway Administration were analyzed for bridges located in Douglas County, which includes the city of Omaha. A total of `r n_bridges` bridges were included in the analysis after excluding records with missing values in key structural or traffic variables.

```{r, echo=FALSE, tbl-summary-stat}
#| tbl-cap: "Descriptive statistics for traffic volume and bridge age among analyzed Douglas County bridges"

summary_table <- tibble(
  Variable = c("Daily Traffic Mean", "Bridge Age (years)"),
  Min = c(adt_min, age_min),
  Median = c(comma(adt_median, age_median)),
  Mean = c(comma(adt_mean, age_mean)),
  Max = c(comma(adt_max, age_max))
)
kable(summary_table, booktabs = TRUE, digits = 1,
      format.args = list(big.mark = ","))

```

As shown in @tbl-summary-stat, Average Daily Traffic ranged from `r comma(adt_min)` to `r comma(adt_max)` vehicles per day, with a median of `r comma(round(adt_median, 0))` and a mean of `r comma(round(adt_mean, 0))`. This wide spread reflects the mix of major arterial crossings and minor local bridges in Douglas County

Bridge age ranged from `r age_min` to `r age_max` years, with a median of `r round(age_median, 1)` years and a mean of `r round(age_mean, 1)` years. The relatively high mean age suggests that a substantial portion of the county's bridge inventory predates modern design standards and may face greater maintenance demands.

The mean deck, superstructure, and substructure ratings were `r round(deck_mean, 2)`, `r round(super_mean, 2)`, and `r round(sub_mean, 2)`, respectively, all falling within the "fair" range (5-6) on the 0-9 NBI scale.


## Structural Condition  

Structural conditions were evaluated using inspection ratings for bridge decks, superstructures, and substructures. These ratings range from 0 (failed condition) to 9 (excellent condition). Since these ratings are recorded as discrete integers rather than continuous measurements, they are best treated as ordered categorical variables; @fig-dist-deck-condition therefore displays counts as a bar chart rather than a histogram.

```{r,echo=FALSE, warning=FALSE}
#| fig-cap: "Distribution of Bridge Deck Condition Ratings in Douglas County, Nebraska."
#| label: fig-dist-deck-condition

cor_age_deck <- cor.test(
  douglas_bridges$bridge_age,
  douglas_bridges$DECK_COND_058,
  method = "spearman"
)

rho_age_deck <- unname(cor_age_deck$estimate)
p_age_deck <- cor_age_deck$p.value

douglas_bridges %>%
  mutate(DECK_COND_058 = factor(DECK_COND_058)) %>%
  ggplot(aes(x = DECK_COND_058,
             fill = DECK_COND_058 %in% as.character(0:4))) +
  geom_bar() +
  scale_fill_manual(values = c("FALSE" = "steelblue", "TRUE" = "darkorange"),
                    labels = c("FALSE" = "Rating ≥ 5", "TRUE" = "Rating ≤ 4 (Higher Risk)"),
                    name   = NULL) +
  labs(
    x = "Deck Condition Rating",
    y = "Number of Bridges"
  ) +
  theme_minimal()
```
The majority of bridges received deck condition ratings of 5, 6, or 7, consistent with the sample mean of `r round(deck_mean, 2)`. However, a meaningful minority of bridges fall at or below the rating-4 threshold. A Spearman correlation test confirmed a statistically significant negative relationship between bridge age and deck condition (ρ = `r round(rho_age_deck, 2)`, p = `r round(p_age_deck, 3)`), indicating that older bridges tend to exhibit lower deck condition ratings — though the effect is modest, suggesting factors beyond age alone (e.g., maintenance history, traffic loading, construction materials) also play a role.

## Traffic Exposure

Traffic exposure was analyzed to evaluate the potential *consequences* of bridge failure. A structurally compromised bridge that carries tens of thousands of vehicles per day poses a fundamentally different public risk than one serving a low-volume rural road.

```{r, echo=FALSE, warning=FALSE}
#| label: fig-adt-distribution
#| fig-cap: "Figure 2. Distribution of Average Daily Traffic (ADT) for bridges in Douglas County, plotted on a logarithmic scale to accommodate the wide range of values. Most bridges carry between 1,000 and 100,000 vehicles per day."

cor_deck_adt <- cor.test(
  douglas_bridges$DECK_COND_058,
  douglas_bridges$ADT_029,
  method = "spearman"
)

rho_deck_adt <- unname(cor_deck_adt$estimate)
p_deck_adt <- cor_deck_adt$p.value

ggplot(douglas_bridges, aes(x = ADT_029)) +
  geom_histogram(bins = 10, fill = "steelblue", color = "white") +
  scale_x_log10(labels = comma) +
  labs(
    x = "Average Daily Traffic (log scale)",
    y = "Number of Bridges"
  ) +
  theme_minimal()
```
As shown in @fig-adt-distribution, the distribution is right-skewed even on a log scale, with a small number of bridges supporting extremely high traffic volumes (exceeding 100,000 vehicles per day). These high-volume bridges represent critical nodes in Omaha's transportation network where a failure would have the most immediate regional impact.

The correlation between deck condition and traffic volume was weak and not statistically significant (ρ = `r round(rho_deck_adt, 2)`, p = `r round(p_deck_adt, 3)`). This is an important finding: bridge condition does not appear to be better or worse on higher-traffic routes, which means high-traffic bridges are not receiving systematically stronger maintenance relative to their exposure.

```{r, echo=FALSE}
risk_summary <- douglas_bridges %>%
  group_by(risk_group) %>%
  summarise(
    n = n(),
    mean_adt = mean(ADT_029, na.rm = TRUE),
    median_adt = median(ADT_029, na.rm = TRUE),
    mean_age = mean(bridge_age, na.rm = TRUE),
    .groups = "drop"
  )

risk_summary
mean_adt_high <- risk_summary %>%
  filter(risk_group == "High Risk") %>%
  pull(mean_adt)

mean_adt_low <- risk_summary %>%
  filter(risk_group == "Lower Risk") %>%
  pull(mean_adt)

n_high <- risk_summary %>%
  filter(risk_group == "High Risk") %>%
  pull(n)

n_low <- risk_summary %>%
  filter(risk_group == "Lower Risk") %>%
  pull(n)
```

Of the `r n_bridges` bridges in the sample, `r n_high` were classified as high risk and `r n_low` as lower risk. High-risk bridges had a mean ADT of `r comma(round(mean_adt_high, 0))`, compared with `r comma(round(mean_adt_low, 0))` for lower-risk bridges.

## Traffic by Bridge Condition

```{r}
#| echo: false
#| fig-cap: "Figure 3. Relationship between bridge structural condition and traffic exposure in Douglas County, Nebraska."

ggplot(douglas_bridges, aes(x = factor(DECK_COND_058), y = ADT_029, fill = factor(DECK_COND_058))) +
  geom_boxplot(fill = "cyan") +
  scale_y_log10() +
  labs(
    x = "Deck Condition Rating (0 = Failed, 9 = Excellent)",
    y = "Average Daily Traffic (Log Scale)"
  ) +
  theme_minimal()
```

The box plot shows that bridges across multiple condition ratings carry a wide range of traffic volumes. While many bridges with moderate condition ratings (around 5–7) carry relatively moderate traffic levels, several bridges with similar condition ratings support substantially higher traffic volumes. The presence of high-traffic outliers indicates that som


::: callout-note

The text seems to have been cut off -- did you forget to push something?

What you have here is closer to a methods section than a results section -- you need to spend more time interpreting your results. This also seems very formulaic -- as if it could have been produced by an AI. I'm not accusing you of that, but I do think you should put a bit more of your personality into your writing style, even in a business report.

What are your actual findings? That's what goes in the results section.

:::
