---
title: "Draft: Results"
editor: 
  markdown: 
    wrap: sentence
---

# Datas & Methods

The analysis uses publicly available bridge inspection and transportation datasets from the **Federal Highway Administration** and summary reports from the **Nebraska Department of Transportation**.
These datasets contain information on bridge conditions, construction characteristics, and traffic volumes for bridges across the United States.
The Nebraska bridge inventory dataset for 2025 was used to evaluate bridge conditions in **Douglas County**, which includes the city of **Omaha**.
The purpose of the analysis is to estimate the potential risk of bridge collapse by examining structural condition ratings and transportation usage.

## Data Sources

-   National Bridge Inventory (NBI): A comprehensive dataset maintained by the Federal Highway Administration that includes inspection results, structural condition ratings, bridge dimensions, construction year, and traffic statistics for bridges in the United States.
-   Nebraska Department of Transportation Bridge Reports: Provides summary information on bridge conditions across Nebraska, including counts of structurally deficient or functionally obsolete bridges and maintenance planning data. These datasets are publicly available and widely used in transportation safety research and infrastructure planning.

## Data Features

The datasets include several variables relevant to evaluating bridge safety and infrastructure risk:

-   Structural Condition Ratings: Includes inspection scores for deck condition `DECK_COND_058`, superstructure condition `SUPERSTRUCTURE_COND_059`, and substructure condition `SUBSTRUCTURE_COND_060`.
    These ratings range from 0 to 9, where higher values represent better structural condition.

-   Traffic Volume: Average Daily Traffic `ADT_029` estimates the number of vehicles crossing each bridge per day, indicating the level of transportation demand.

-   Bridge Age: Construction year `YEAR_BUILT_027` allows calculation of bridge age, which may influence structural deterioration over time.

-   Location Information: Geographic identifiers such as state code and county code allow bridges in Douglas County to be isolated for analysis.

-   Bridge Dimensions and Characteristics: Additional variables such as bridge length, roadway width, and number of spans describe the structural characteristics of each bridge

## Data Collection and Preparation

The Nebraska National Bridge Inventory dataset was downloaded in delimited text format and converted into a CSV file for analysis.
The dataset was then imported into the R programming environment.

To focus on the study area, bridges located in Douglas County were selected using the county code variable.
Data cleaning steps were performed to prepare the dataset for statistical analysis.
Several inspection variables were initially stored as character values and were converted to numeric format in order to calculate summary statistics.
Missing or non-numeric values were treated as missing observations.

Additional variables were created to support analysis.
Bridge age was calculated by subtracting the construction year from the current year.
This variable helps identify older infrastructure that may require increased maintenance or monitoring.

## Data Classification

To evaluate structural risk, bridges were categorized according to their inspection ratings.

-   Higher-risk bridges: Bridges with structural condition ratings of 4 or lower in the deck, superstructure, or substructure categories.
    These ratings indicate poor structural condition or advanced deterioration.

-   Moderate-condition bridges: Bridges with ratings between 5 and 6, indicating fair structural condition.

-   Good-condition bridges: Bridges with ratings of 7 or higher, representing satisfactory or good structural performance.

This classification allows the analysis to identify bridges that may require additional inspection, repair, or replacement planning.

## R packages and Software

The analysis was conducted in R using several packages for data management and visualization:

-   **readr:** Used to import the bridge inventory dataset into R.

-   **tidyverse:** Used for data manipulation, filtering, and summarization.

-   **dplyr:** Used to calculate descriptive statistics and identify high-risk bridges.

-   **ggplot2:** Used to create visualizations of bridge condition ratings, traffic levels, and bridge age distributions.

-   **knitr:** Used to generate the final report and integrate code with written analysis.

## Analytical Approach

To estimate bridge collapse risk and identify vulnerable structures, several statistical analyses were conducted:

-   Descriptive statistics were calculated to summarize the number of bridges, average structural condition ratings, bridge age, and traffic volume within Douglas County.

-   Condition rating analysis was performed to identify bridges with structural condition ratings indicating potential deterioration.

-   Traffic exposure analysis examined bridges with high average daily traffic to assess the potential consequences of structural failure.

-   Combined risk assessment identified bridges that simultaneously exhibit lower structural condition ratings and high traffic volumes.
    These bridges represent the greatest infrastructure risk because they combine higher likelihood of structural problems with greater transportation impact.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(readr)
library(tidyverse)
library(dplyr)
library(ggplot2)
```

```{r,echo=FALSE, warning=FALSE}
inventory <- read.csv("data/national_bridge_inventory_ne.csv")

bridges_clean <- inventory %>%
  select(
    COUNTY_CODE_003,
    YEAR_BUILT_027,
    YEAR_RECONSTRUCTED_106,
    ADT_029,
    FUNCTIONAL_CLASS_026,
    DECK_COND_058,
    SUPERSTRUCTURE_COND_059,
    SUBSTRUCTURE_COND_060,
    LOWEST_RATING,
    BRIDGE_CONDITION,
    STRUCTURE_LEN_MT_049,
    DECK_AREA
  )

douglas_bridges <- bridges_clean %>%
  filter(COUNTY_CODE_003 == 55) %>%
  filter(
    !is.na(DECK_COND_058),
    !is.na(SUPERSTRUCTURE_COND_059),
    !is.na(SUBSTRUCTURE_COND_060)
  )

#bridge age
douglas_bridges <- douglas_bridges %>%
  mutate(bridge_age = 2026 - YEAR_BUILT_027)

douglas_bridges <- douglas_bridges %>%
  mutate(
    DECK_COND_058 = as.numeric(DECK_COND_058),
    SUPERSTRUCTURE_COND_059 = as.numeric(SUPERSTRUCTURE_COND_059),
    SUBSTRUCTURE_COND_060 = as.numeric(SUBSTRUCTURE_COND_060)) %>%
  drop_na(DECK_COND_058, SUPERSTRUCTURE_COND_059, SUBSTRUCTURE_COND_060)



high_risk <- douglas_bridges %>%
  filter(
    DECK_COND_058 <= 4 |
    SUPERSTRUCTURE_COND_059 <= 4 |
    SUBSTRUCTURE_COND_060 <= 4
  )



```

# Results

## Study Sample Characteristics

Bridge inspection data from the Federal Highway Administration were analyzed for bridges located in Douglas County, which includes the city of Omaha.
A total of 8 bridges were included in the dataset used for analysis.

Traffic volumes vary substantially across bridges.
Average Daily Traffic (ADT) ranges from 25 vehicles per day to 182,320 vehicles per day, with a median of 15,750 vehicles per day and an average of 36,947 vehicles per day.
Bridge age ranges from 2 to 106 years, with a median age of 39 years.

Table 1 summarizes the key descriptive statistics for traffic volume and bridge age.

| Variable              | Min | Median | Mean   | Max     |
|-----------------------|-----|--------|--------|---------|
| Average Daily Traffic | 25  | 15,750 | 36,947 | 182,320 |
| Bridge Age (years)    | 2   | 39     | 41     | 106     |

## Structural Condition  

Structural conditions were evaluated using inspection ratings for bridge decks, superstructures, and substructures. These ratings range from 0 (failed condition) to 9 (excellent condition).

Across the bridges in the sample, the average structural condition ratings are approximately 6, indicating that most bridges fall within the fair condition category. Figure 1 illustrates the distribution of deck condition ratings across bridges.

```{r,echo=FALSE, warning=FALSE}
#| fig-cap: "Figure 1. Distribution of Bridge Deck Condition Ratings in Douglas County, Nebraska."

douglas_bridges %>%
  summarise(
    avg_deck = mean(DECK_COND_058, na.rm = TRUE),
    avg_super = mean(SUPERSTRUCTURE_COND_059, na.rm = TRUE),
    avg_sub = mean(SUBSTRUCTURE_COND_060, na.rm = TRUE)
  )

ggplot(douglas_bridges, aes(x = DECK_COND_058)) +
  geom_histogram(binwidth = 1) +
  labs(
    title = "Distribution of Bridge Deck Condition Ratings",
    x = "Deck Condition Rating",
    y = "Number of Bridges"
  )
```


## Traffic Exposure

Traffic exposure was analyzed to evaluate the potential consequences of bridge failure. Bridges with higher traffic volumes represent greater transportation risk because a structural failure could disrupt a larger number of vehicles and commuters.

The distribution of Average Daily Traffic across bridges is shown in Figure 2. While several bridges carry relatively low traffic volumes, others support substantially higher traffic levels, indicating their importance within the regional transportation network.

```{r, echo=FALSE, warning=FALSE}
#| fig-cap: "Figure 2. Distribution of Average Daily Traffic (ADT) for bridges in Douglas County."

ggplot(douglas_bridges, aes(x = ADT_029)) +
  geom_histogram(bins = 10) +
  scale_x_log10() +
  labs(
    x = "Average Daily Traffic (log scale)",
    y = "Number of Bridges"
  ) +
  theme_minimal()

```

Figure 2 shows the distribution of Average Daily Traffic (ADT) for bridges in Douglas County using a logarithmic scale. The use of a log scale allows bridges with very different traffic volumes to be visualized more clearly.

The distribution indicates that most bridges carry moderate to high levels of traffic, with many bridges falling between 10,000 and 100,000 vehicles per day. A smaller number of bridges carry relatively low traffic volumes, while a few bridges support extremely high traffic levels exceeding 100,000 vehicles per day.

## Traffic by Bridge Condition
```{r}
#| echo: false
#| fig-cap: "Figure 3. Relationship between bridge structural condition and traffic exposure in Douglas County, Nebraska."

ggplot(douglas_bridges, aes(x = factor(DECK_COND_058), y = ADT_029, fill = factor(DECK_COND_058))) +
  geom_boxplot(fill = "cyan") +
  scale_y_log10() +
  labs(
    x = "Deck Condition Rating (0 = Failed, 9 = Excellent)",
    y = "Average Daily Traffic (Log Scale)",
    fill = "Condition Rating"
  ) +
  theme_minimal()
```
The box plot shows that bridges across multiple condition ratings carry a wide range of traffic volumes. While many bridges with moderate condition ratings (around 5–7) carry relatively moderate traffic levels, several bridges with similar condition ratings support substantially higher traffic volumes. The presence of high-traffic outliers indicates that som