---
title: "Result"
format: html
---


```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(readr)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(knitr)
library(scales)

```

```{r,echo=FALSE, warning=FALSE}
inventory <- read.csv("data/national_bridge_inventory_ne.csv")

douglas_bridges <- inventory %>%
  select(
    COUNTY_CODE_003,
    YEAR_BUILT_027,
    YEAR_RECONSTRUCTED_106,
    ADT_029,
    FUNCTIONAL_CLASS_026,
    DECK_COND_058,
    SUPERSTRUCTURE_COND_059,
    SUBSTRUCTURE_COND_060,
    LOWEST_RATING,
    BRIDGE_CONDITION,
    STRUCTURE_LEN_MT_049,
    DECK_AREA
  ) %>%
  filter(COUNTY_CODE_003 == 55) %>%
  mutate(
    YEAR_BUILT_027 = as.numeric(YEAR_BUILT_027),
    ADT_029 = as.numeric(ADT_029),
    DECK_COND_058 = as.numeric(DECK_COND_058),
    SUPERSTRUCTURE_COND_059 = as.numeric(SUPERSTRUCTURE_COND_059),
    SUBSTRUCTURE_COND_060 = as.numeric(SUBSTRUCTURE_COND_060),
    bridge_age = 2026 - YEAR_BUILT_027
  ) %>%
  drop_na(ADT_029, bridge_age, DECK_COND_058, SUPERSTRUCTURE_COND_059, SUBSTRUCTURE_COND_060)

douglas_bridges <- douglas_bridges %>%
  mutate(
    risk_group = ifelse(
      DECK_COND_058 <= 4 |
      SUPERSTRUCTURE_COND_059 <= 4 |
      SUBSTRUCTURE_COND_060 <= 4,
      "High Risk",
      "Lower Risk"
    )
  )
```


```{r, echo=FALSE}
n_bridges <- nrow(douglas_bridges)

adt_min <- min(douglas_bridges$ADT_029, na.rm = TRUE)
adt_max <- max(douglas_bridges$ADT_029, na.rm = TRUE)
adt_median <- median(douglas_bridges$ADT_029, na.rm = TRUE)
adt_mean <- mean(douglas_bridges$ADT_029, na.rm = TRUE)

age_min <- min(douglas_bridges$bridge_age, na.rm = TRUE)
age_max <- max(douglas_bridges$bridge_age, na.rm = TRUE)
age_median <- median(douglas_bridges$bridge_age, na.rm = TRUE)
age_mean <- mean(douglas_bridges$bridge_age, na.rm = TRUE)

deck_mean <- mean(douglas_bridges$DECK_COND_058, na.rm = TRUE)
super_mean <- mean(douglas_bridges$SUPERSTRUCTURE_COND_059, na.rm = TRUE)
sub_mean <- mean(douglas_bridges$SUBSTRUCTURE_COND_060, na.rm = TRUE)

```

## Study Sample Characteristics

Bridge inspection data from the Federal Highway Administration were analyzed for bridges located in Douglas County, which includes the city of Omaha. A total of `r n_bridges` bridges were included in the analysis after excluding records with missing values in key structural or traffic variables.

```{r, echo=FALSE, tbl-summary-stat}
#| tbl-cap: "Descriptive statistics for traffic volume and bridge age among analyzed Douglas County bridges"

summary_table <- tibble(
  Variable = c("Daily Traffic Mean", "Bridge Age (years)"),
  Min = c(adt_min, age_min),
  Median = c(comma(adt_median),comma(age_median)),
  Mean = c(comma(adt_mean),comma(age_mean)),
  Max = c(comma(adt_max),comma(age_max))
)
kable(summary_table, booktabs = TRUE, digits = 1,
      format.args = list(big.mark = ","))

```

As shown in @tbl-summary-stat, Average Daily Traffic ranged from `r comma(adt_min)` to `r comma(adt_max)` vehicles per day, with a median of `r comma(round(adt_median, 0))` and a mean of `r comma(round(adt_mean, 0))`. This wide spread reflects the mix of major arterial crossings and minor local bridges in Douglas County

Bridge age ranged from `r age_min` to `r age_max` years, with a median of `r round(age_median, 1)` years and a mean of `r round(age_mean, 1)` years. The relatively high mean age suggests that a substantial portion of the county's bridge inventory predates modern design standards and may face greater maintenance demands.

The mean deck, superstructure, and substructure ratings were `r round(deck_mean, 2)`, `r round(super_mean, 2)`, and `r round(sub_mean, 2)`, respectively, all falling within the "fair" range (5-6) on the 0-9 NBI scale, while the superstructure mean marginally exceeds the good threshold ($\geq$ 7), indicating the overall structural inventory is in fair-to good condition.


## Structural Condition  

Structural conditions were evaluated using inspection ratings for bridge decks, superstructures, and substructures. These ratings range from 0 (failed condition) to 9 (excellent condition). Since these ratings are recorded as discrete integers rather than continuous measurements, they are best treated as ordered categorical variables; @fig-dist-deck-condition therefore displays counts as a bar chart rather than a histogram.

```{r,echo=FALSE, warning=FALSE}
#| fig-cap: "Distribution of Bridge Deck Condition Ratings in Douglas County, Nebraska."
#| label: fig-dist-deck-condition

cor_age_deck <- cor.test(
  douglas_bridges$bridge_age,
  douglas_bridges$DECK_COND_058,
  method = "spearman"
)

rho_age_deck <- unname(cor_age_deck$estimate)
p_age_deck <- cor_age_deck$p.value

douglas_bridges %>%
  mutate(DECK_COND_058 = factor(DECK_COND_058)) %>%
  ggplot(aes(x = DECK_COND_058,
             fill = DECK_COND_058 %in% as.character(0:4))) +
  geom_bar() +
  scale_fill_manual(values = c("FALSE" = "steelblue", "TRUE" = "darkorange"),
                    labels = c("FALSE" = "Rating ≥ 5", "TRUE" = "Rating ≤ 4 (Higher Risk)"),
                    name   = NULL) +
  labs(
    x = "Deck Condition Rating",
    y = "Number of Bridges"
  ) +
  theme_minimal()
```
The majority of bridges received deck condition ratings of 5, 6, or 7, consistent with the sample mean of `r round(deck_mean, 2)`. However, a meaningful minority of bridges fall at or below the rating-4 threshold. A Spearman correlation test confirmed a statistically significant negative relationship between bridge age and deck condition ($\rho$ = `r round(rho_age_deck, 2)`, p = `r round(p_age_deck, 3)`), indicating that older bridges tend to exhibit lower deck condition ratings — though the effect is modest, suggesting factors beyond age alone (e.g., maintenance history, traffic loading, construction materials) also play a role.

## Traffic Exposure

Traffic exposure was analyzed to evaluate the potential *consequences* of bridge failure. A structurally compromised bridge that carries tens of thousands of vehicles per day poses a fundamentally different public risk than one serving a low-volume rural road.

```{r, echo=FALSE, warning=FALSE, fig.height=4, fig.width=6}
#| label: fig-adt-distribution
#| fig-cap: "Distribution of Average Daily Traffic (ADT) for bridges in Douglas County, plotted on a logarithmic scale to accommodate the wide range of values. Most bridges carry between 1,000 and 100,000 vehicles per day."

cor_deck_adt <- cor.test(
  douglas_bridges$DECK_COND_058,
  douglas_bridges$ADT_029,
  method = "spearman"
)

rho_deck_adt <- unname(cor_deck_adt$estimate)
p_deck_adt <- cor_deck_adt$p.value

ggplot(douglas_bridges, aes(x = ADT_029)) +
  geom_histogram(bins = 10, fill = "steelblue", color = "white") +
  scale_x_log10(labels = comma) +
  labs(
    x = "Average Daily Traffic (log scale)",
    y = "Number of Bridges"
  ) +
  theme_minimal()
```

As shown in @fig-adt-distribution, the distribution is left-skewed even on a log scale, with a small number of bridges supporting extremely high traffic volumes (exceeding 100,000 vehicles per day). These high-volume bridges represent critical nodes in Omaha's transportation network where a failure would have the most immediate regional impact.

The correlation between deck condition and traffic volume was weak and not statistically significant ($\rho$ = `r round(rho_deck_adt, 2)`, p = `r round(p_deck_adt, 3)`). This is an important finding: bridge condition does not appear to be better or worse on higher-traffic routes, which means high-traffic bridges are not receiving systematically stronger maintenance relative to their exposure.

```{r, echo=FALSE}
risk_summary <- douglas_bridges %>%
  group_by(risk_group) %>%
  summarise(
    n = n(),
    mean_adt = mean(ADT_029, na.rm = TRUE),
    median_adt = median(ADT_029, na.rm = TRUE),
    mean_age = mean(bridge_age, na.rm = TRUE),
    .groups = "drop"
  )

mean_adt_high <- risk_summary %>%
  filter(risk_group == "High Risk") %>%
  pull(mean_adt)

mean_adt_low <- risk_summary %>%
  filter(risk_group == "Lower Risk") %>%
  pull(mean_adt)

mean_age_high <- risk_summary %>% 
  filter(risk_group == "High Risk")  %>% 
  pull(mean_age)

mean_age_low  <- risk_summary %>% 
  filter(risk_group == "Lower Risk") %>% 
  pull(mean_age)

n_high <- risk_summary %>%
  filter(risk_group == "High Risk") %>%
  pull(n)

n_low <- risk_summary %>%
  filter(risk_group == "Lower Risk") %>%
  pull(n)
```

Of the `r n_bridges` bridges analyzed, `r n_high` (`r round(100 * n_high / n_bridges, 1)`%) were classified as higher-risk based on having at least one structural condition rating at or below 4. The remaining `r n_low` bridges were classified as lower risk.

Higher-risk bridges were, on average, `r round(mean_age_high - mean_age_low, 1)` years older than lower-risk bridges (mean age `r round(mean_age_high, 1)` vs. `r round(mean_age_low, 1)` years), reinforcing the age–condition relationship observed above. Their mean daily traffic (`r comma(round(mean_adt_high, 0))` vehicles/day) was `r ifelse(mean_adt_high > mean_adt_low, "higher", "lower")` than that of lower-risk bridges (`r comma(round(mean_adt_low, 0))` vehicles/day), though the difference should be interpreted cautiously given the high variability in ADT across both groups.

## Traffic by Bridge Condition

The combined risk picture — structural condition against traffic load — is the central question for prioritizing interventions. @fig-condition-traffic displays the distribution of ADT by deck condition rating.

```{r}
#| echo: false
#| label: fig-condition-traffic
#| fig-cap: "Relationship between bridge structural condition and traffic exposure in Douglas County, Nebraska."

ggplot(douglas_bridges,
       aes(x = factor(DECK_COND_058), y = ADT_029,
           fill = DECK_COND_058 <= 4)) +
  geom_boxplot() +
  scale_y_log10(labels = comma) +
  scale_fill_manual(values = c("FALSE" = "steelblue", "TRUE" = "darkorange"),
                    labels = c("FALSE" = "Rating ≥ 5", "TRUE"  = "Rating ≤ 4 (Higher Risk)"),
                    name   = NULL) +
  labs(
    x = "Deck Condition Rating (0 = Failed, 9 = Excellent)",
    y = "Average Daily Traffic (log scale)"
  ) +
  theme_minimal()
```

@fig-condition-traffic reveals a finding that carries direct policy implications: several bridges rated at 4 or below carry daily traffic volumes in the tens of thousands. These bridges sit in the upper-left danger zone — structurally compromised yet heavily used — and represent the clearest candidates for prioritized inspection and intervention. At the same time, the wide interquartile ranges within each rating category confirm that condition rating alone is not a reliable proxy for traffic exposure; both dimensions must be considered together when allocating limited maintenance resources.


